Sara Benist MADA Project

### 2023-04-07 code/analysis_code/statistical_analysis.qmd
Part 4a of project
machine learning analysis

---
title: "Statistical Analysis"
author: "Sara Benist"
date: "2023-03-17"
output: html_document
---

@@

This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script. The processed data is used for model fitting.

@@

```{r, loaddata}
```
# Model fitting
Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set. First, we will focus on linear models of the outcomes TB incidence, TB mortality, and TB prevalence. I will be fitting two models using the full data set and only the high burden countries.

## Linear model fitting for TB Incidence

```{r, fitmodel1}
```
The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1. The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`. Since many of the countries in the full data set have very low incidence, the coefficients for the intercept and subgroup variable are much lower. The model has more data points with low values.

### Prediction model for TB Incidence
We can also try to predict the approximate TB incidence for a group. For this example, we will include subgroup and setting as predictors.  
```{r, pred1}
```
With the prediction model including setting, there are a lot of variables in the model. The prediction would have a 1 for the country of interest and 0 for all other setting variables. For example, the prediction for a male in Central African Republic would appear as `TB incidence = 126(1)+189(1)`.

As an example, I will predict the incidence rate for males and females in India. The code for prediction modeling is based off of the `tidymodels` guidance. The plot shows the incidence rate and confidence intervals for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) in India. This can be repeated with the high burden countries.

## Linear model fitting for TB Mortality

```{r, fitmodel2}
```
For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`.

## Linear model fitting for TB Prevalence

```{r, fitmodel3}
```
For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0. The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`. The difference between the high burden and full data models is lower than the other outcomes, most likely due to the long duration of TB or less data points available with the full data model.

## Linear model fitting for BCG Coverage
Now we can look at some of the other indicators in the data set. I will only be looking at the indicators that had variation between subgroups. 
```{r, fitmodel4}
```
The BCG coverage has 12 variables that produce a linear equation of `BCG coverage % = 88.4 + 0.363(Male) -10.6(no education) + 0.169(primary education) + 6.66(higher education) - 6.46(quintile 1) - 1.6(quintile 2) +1.06(quintile 3) + 4.50(quintile 4) + 7.01(quintile 5) - 2.12(rural) + 5.10(Urban)`. The full model has an equation of `BCG coverage % = 88.3 + 0.245(Male) -9.44(no education) + 0.958(primary education) + 6.42(higher education) - 6.40(quintile 1) - 1.68(quintile 2) + 0.773(quintile 3) + 3.84(quintile 4) + 6.41(quintile 5) - 2.08(rural) + 4.75(Urban)`. The two models are overall pretty similar and can be used in the same way as the prediction model (entering in 1 or 0 depending on categorization).

## Linear model fitting for Catastrophic costs

```{r, fitmodel5}
```
The high burden model has an equation of `% catastrophic costs = 87.3 - 40.4(drug-suseptible TB) - 18.6(quintile 1) - 38.8(quintile 2) - 44.6(quintile 3) - 47.4(quintile 4) - 58.5(quintile 5)`. The full model data has an equation of `% catastrophic costs = 83.5 - 35.1(drug-suseptible TB) - 11.1(quintile 1) - 30.9(quintile 2) - 38.4(quintile 3) - 41.4(quintile 4) - 54.9(quintile 5)`.
## Linear model fitting for case detection rate

```{r, fitmodel6}
```
For high burden countries, the equation is `case detection rate % = 37.9 + 19.1(15+ or older) + 16.7(female) + 16.4(male)`. The full data model has an equation of `case detection rate % = 43.8 + 18.9(15+ or older) + 15.3(female) + 16.9(male)`.

## Linear model fitting for prevalence to notification ratio

```{r, fitmodel7}
```
The high burden model has an equation of `prevalence to notification ratio in years = 1.73 + 0.664(male)`, and the full model has an equation of `prevalence to notification ratio in years = 1.73 + 0.701(male)`.

## Linear model fitting for attitudes

```{r, fitmodel8}
```
The attitude question regarding TB was about the desire to keep a diagnosis hidden. Since males are more likely to be diagnosed with TB, I focused on the male attitude and knowledge. For high burden countries, the equation is `% wanting to keep TB secret = 27.6 - 5.27(20-34 years) - 8.21(35-49 years) - 3.53(no education) - 2.71(primary education) - 7.23(higher education) - 5.01(quintile 1) - 5.23(quintile 2) - 4.97(quintile 3) - 4.38(quintile 4) - 5.86(quintile 5) - 4.91(rural) - 6.27(urban)`. For the full model, the equation is `% wanting to keep TB secret = 24.7 - 4.94(20-34 years) - 8.14(35-49 years) - 4.32(no education) - 3.08(primary education) - 6.41(higher education) - 4.44(quintile 1) - 5.24(quintile 2) - 5.18(quintile 3) - 4.59(quintile 4) - 6.20(quintile 5) - 4.66(rural) - 5.99(urban)`.

## Linear model fitting for knowledge

```{r, fitmodel9}
```
The knowledge question asked about if TB is spread through cough. We will again focus on male knowledge.

The high burden countries had an equation of `% responding TB is spread through cough = 65.9 + 4.83(20-34 years) + 6.19(35-49 years) - 9.91(no education) - 2.81(primary education) + 12.6(higher education) - 6.01(quintile 1) - 0.464(quintile 2) + 2.32(quintile 3) + 7.50(quintile 4) + 14.6(quintile 5) + 0.419(rural) + 11.8(urban)`. The full model had an equation of `% responding TB is spread through cough = 64.6 + 6.17(20-34 years) + 7.29(35-49 years) - 8.55(no education) - 1.66(primary education) + 12.8(higher education) - 3.88(quintile 1) + 0.028(quintile 2) + 3.03(quintile 3) + 8.18(quintile 4) + 15.8(quintile 5) + 1.28(rural) + 12.8(urban)`.

I would like to try and create a dataset using the high burden countries in order to predict the main outcomes of TB incidence and mortality from the indicators (BCG Coverage, knowledge, attitudes, etc). Right now, the data is not set up to complete this analysis, but I will try to find a way to do this.

>>>>

---
title: "Statistical Analysis"
author: "Sara Benist"
date: "2023-04-06"
output: html_document
---

@@

This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script. The best way to view the output of the script is to render the qmd and view the html document. The processed data is used for model fitting.

@@

```{r, loaddata}
```
# Model fitting
Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set. First, we will focus split the data into testing and training sets and set up the cross-validation folds. The training set will be used to tune the model parameters and the testing set will be used to test model performance on new data.

```{r, splitdata}

```

I will be tuning decision tree models for the main outcomes of TB incidence, TB mortality, and TB prevalence for both high burden countries and the full data set.

## Decision tree model for TB Incidence

```{r, wdincidencemodel}
```
The best tree model for TB incidence using the full data has a depth of 4 with an RMSE of 128. This is a very low level of performance in predicting TB incidence. The fitted model to the training data has a RMSE of 62.8, but using the testing data, the model had a performance of RMSE = 110. The most important predictor was country.

```{r, hbincidencemodel}

```
The best tree model for TB incidence using the high burden countries has a depth of 4 with an RMSE of 162. This is also a low level of performance in predicting TB incidence. The fitted model to the training data has a RMSE of 126 which was much higher compared to the full dataset. Using the testing data, the model had a performance of RMSE = 134. The most important predictor was also country.
## Decision tree model for TB Mortality

```{r}
#recipe predicting TB mortality
mortrecipe <- recipe(`TB mortality (deaths per 100 000 population)` ~.,
                 data = wdtrain_data)
#create tree model
treemodel <- decision_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
) %>% 
  set_engine("rpart") %>% 
  set_mode("regression")

#create workflow
mortwflow <- workflow() %>% 
  add_model(treemodel) %>% 
  add_recipe(mortrecipe)

#set seed for reproducibility
set.seed(626)
#resample using cross-validation folds to tune parameters
morttree_resamp <- mortwflow %>% 
  tune_grid(
    resamples = wdfolds,
    grid = tree_grid,
    control = control_grid(save_pred = TRUE)
  )

#plot resampling by parameters
morttreeplot <- morttree_resamp %>% autoplot()
morttreeplot

#show the best model
morttree_resamp %>% 
  show_best()

#select best model
wdmortbest_tree <- morttree_resamp %>% select_best("rmse")

#create final workflow
wdmortfinal_wf <- mortwflow %>% finalize_workflow(wdmortbest_tree)

#fit the final workflow to the training data
wdmortfinal_fit <- wdmortfinal_wf %>% fit(wdtrain_data)

#find RMSE for model with training data
wdmortfitted <- augment(wdmortfinal_fit, wdtrain_data) %>% 
  select(`TB mortality (deaths per 100 000 population)`, .pred) %>% 
  rmse(truth = `TB mortality (deaths per 100 000 population)`, .pred)
wdmortfitted
#fit final workflow to test data
wdmortfittest <- wdmortfinal_fit %>% last_fit(wddatasplit)

#find RMSE for model with test data
wdmortfittest %>% collect_metrics()

#check important variables of model
wdmortfittest %>% extract_fit_parsnip() %>% vip()

```
The best tree model for TB mortality using the full data has a depth of 1 with an RMSE of 16.9. The fitted model to the training data has a RMSE of 11.3, but using the testing data, the model had a performance of RMSE = 17.4. The most important predictor was country.

```{r, hbmortalitymodel}
```
The best tree model for TB mortality using the high burden countries has a depth of 1 with an RMSE of 20.8. The fitted model to the training data has a RMSE of 16.0. Using the testing data, the model had a performance of RMSE = 21.0. The most important predictor was country.

## Decision tree model for TB Prevalence

```{r, wdprevalencemodel}
```
The best tree model for TB prevalence using the full data has a depth of 1 with an RMSE of 208. This is a very low level of performance in predicting TB prevalence. The fitted model to the training data has a RMSE of 135, and using the testing data, the model had a performance of RMSE = 287. The most important predictor was country.

```{r, hbprevalencemodel}
```
The best tree model for TB prevalence using the high burden countries has a depth of 1 with an RMSE of 172. This is a very low level of performance in predicting TB incidence. The fitted model to the training data has a RMSE of 138, but using the testing data, the model had a performance of RMSE = 237. The most important predictor was country.

The models were overall not a good predictor of TB outcomes. This is possibly due to the large variance in the outcomes as well as the low number of predictors available. Since country is consistently the most important predictor, we can look at some of the health indicators to see some of the underlying trends that can occur within a country.

## Decision tree model for BCG coverage

```{r, wdbcgmodel}
```
The best tree model for BCG coverage using the full data has a depth of 15 with an RMSE of 5.86. This would appear to be a better prediction than the previous ones. The fitted model to the training data has a RMSE of 4.29, and using the testing data, the model had a performance of RMSE = 5.30. The most important predictor was country followed by subgroup.

```{r, hbbcgmodel}
```
The best tree model for BCG coverage using high burden countries has a depth of 11 with an RMSE of 5.79. The fitted model to the training data has a RMSE of 4.00, but using the testing data, the model had a performance of RMSE = 5.56. These are approximately similar to the full data set. The most important predictor was country.

## Decision tree model for Catastrophic Cost
```{r, wdcatacostmodel}
```
The best tree model for catastrophic costs using the full data has a depth of 4 with an RMSE of 17.2. The fitted model to the training data has a RMSE of 10.2, but using the testing data, the model had a performance of RMSE = 18.1. The most important predictor was subgroup.

```{r, hbcatacostmodel}
```
The best tree model for catastrophic costs using the high burden countries has a depth of 4 with an RMSE of 16.3. The fitted model to the training data has a RMSE of 10.7, but using the testing data, the model had a performance of RMSE = 15.2. The most important predictor was subgroup.

## Decision tree model for Case Detection Rate

```{r, wdcasedetectmodel}
```
The best tree model for case detection rate using the full data has a depth of 8 with an RMSE of 13.6. The fitted model to the training data has a RMSE of ###, but using the testing data, the model had a performance of RMSE = 12.2. The most important predictor was country.

```{r, hbcasedetectmodel}
```
The best tree model for case detection rate using high burden countries has a depth of 8 with an RMSE of 16.1. This is a very low level of performance in predicting TB incidence. The fitted model to the training data has a RMSE of 7.93, but using the testing data, the model had a performance of RMSE = 11.8. The most important predictor was country.

## Decision tree model for Prevalence to Notification ratio

```{r, wdptnmodel}
```
The best tree model for prevalence to notification ratio using the full data has a depth of 4 with an RMSE of 1.06. This model has the highest performance so far. The fitted model to the training data has a RMSE of 0.546, and using the testing data, the model had a performance of RMSE = 0.656. The most important predictor was country.

```{r, hbptnmodel}
```
The best tree model for prevalence to notification ratio using high burden countries has a depth of 1 with an RMSE of 0.984. This model also has high performance and has a different tree depth than the full data set. The fitted model to the training data has a RMSE of 0.650, but using the testing data, the model had a performance of RMSE = 0.773. The most important and only predictor was country.

## Decision tree model for Male Attitudes

```{r, wdmaleattitudemodel}
```
The best tree model for male attitudes using the full data has a depth of 4 with an RMSE of 5.39. The fitted model to the training data has a RMSE of 3.87, and using the testing data, the model had a performance of RMSE = 4.62. The most important predictor was country.

```{r, hbmaleattitubemodel}
```
The best tree model for male attitude using high burden countries has a depth of 8 with an RMSE of 5.70. The fitted model to the training data has a RMSE of 3.71, but using the testing data, the model had a performance of RMSE = 4.46. The most important predictor was country.

## Decision tree model for Male Knowledge
```{r, wdmaleknowledgemodel}
```
The best tree model for male knowledge using the full data has a depth of 8 with an RMSE of 8.52. The fitted model to the training data has a RMSE of 5.69, but using the testing data, the model had a performance of RMSE = 6.49. The most important predictor was country.

```{r, hbmaleknowledgemodel}
```
The best tree model for male knowledge using the high burden countries has a depth of 4 with an RMSE of 10.8. The fitted model to the training data has a RMSE of 6.23, but using the testing data, the model had a performance of RMSE = 8.99. The most important predictor was year which was different from the full data set.

### 2023-04-07 products/manuscript/Manuscript.qmd
Part 4a of project
machine learning analysis

TBA: background of why and how I am studying this topic

@@

*(Part 1 of project: data source, description, loading into raw folder, general aspects)*

The World Health Organization collected data on the inequity surrounding tuberculosis, HIV, and malaria for the [State of inequality report](https://www.who.int/data/inequality-monitor/publications/report_2021_hiv_tb_malaria), and I will be exploring the dataset for TB. More information about the data can be found [here](https://heatrepository.blob.core.windows.net/documents/data-repository-indicator-list.pdf?sp=r&st=2022-06-07T14:16:36Z&se=2023-12-30T23:16:36Z&spr=https&sv=2020-08-04&sr=b&sig=4kzThU1QDo55UOQyhWcUd8rPWJ9LxUZdRRI3zl6wKNs%3D) under "Tuberculosis Indicators". The dataset can be found [here](o%20https://www.who.int/data/inequality-monitor/data#PageContent_C158_Col00) under "Tuberculosis Indicators".

For the tuberculosis dataset, the data was collected from the WHO Global TB programme, TB prevalence surveys, country-specific TB programmes, the WHO Health Equity Monitor database, TB patient cost surveys, and other sources. The WHO organized the dataset to be used with the Health Equity Assessment Toolkit which is the built in data analysis and exploration tool. The database contains 10 variables regarding the burden, detection, prevention, knowledge, and social protection and observations for 194 countries over various years. The observations for each variable is further separated by up 7 inequality domains. Not all countries have data available for each year or for each inequality domain. The dataset contains a total of 7473 observations.

@@

*(Part 1 of project: RQ, outcomes, predictors, patterns)*

Research question: How do the TB indicators relate to the TB incidence, prevalence, and mortality of a country, and what inequality measure shows the greatest disparity in TB outcomes?

@@

*(Part 1 of project: initial analysis thoughts)*

@@

## Data aquisition, import, and cleaning
*(Part 2 of project: data importing and cleaning)*

The WHO dataset can be found [here](o%20https://www.who.int/data/inequality-monitor/data#PageContent_C158_Col00) under "Tuberculosis Indicators". This data was downloaded and stored into the `raw_data` folder of the repository as `rawdata`. Information on the indicators, social determinants dimensions, and subgroups can be found in the `raw_data` folder under `202206-metadata-tb.pdf`
```{r load data}
#path to data
data_location <- here::here("data","raw_data","202206-repository-tb.xlsx")

@@

str(rawdata)
```
Please see `processingfile.qmd` for full code describing data importing and cleaning.

@@

For the main part of the data cleaning, I used `pivot_wider()` to allow each indicator to have its own column, and subset the high burden countries (as indicated by the WHO) into a new object called `highburden`. I also checked each indicator using `filter()`, `select()`, and `summary()` functions. These methods will allow for deeper exploration of each indicator in order to explore trends and patterns for each subgroup. An example of the data cleaning is shown below. 
```{r}
wide_data <- rawdata %>% 
  select(c(setting, year, indicator_name,
           indicator_abbr, dimension, subgroup,
           estimate, population)) %>% 
  pivot_wider(names_from = "indicator_name", values_from = "estimate")
head(wide_data)

@@

The processed and cleaned data was then saved into the `processed_data` folder under `processeddata.rda` to be used during the exploratory analysis phase. I have loaded the processed data below.
```{r}
#path to data
data_location <- here::here("data","processed_data","processeddata.rda")
#load data. 
load(data_location)
```

## Statistical analysis
*(Part 3 of project: statistical analysis)*
For the statistical analysis of the data, I created linear models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator. An example is shown below. This method was repeated for the `wide_data` and `highburden` countries.

@@

Other indicators were also analyzed and fit to a linear model as shown below. 
```{r}
# fit linear model with BCG coverage as outcome, subgroups as predictors

@@

## Exploratory/Descriptive analysis
*(Part 2 of project: exploratory analysis)*
Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.

@@

#mortality summary and save to file location
summary_mort <- skimr::skim(wide_data$`TB mortality (deaths per 100 000 population)`)
print(summary_mort)
```
The population for most variables were relatively low and skewed to the right. The TB incidence and mortality indicators also showed a skewed distribution. The highest incidence rate was 908 cases/100,000 and highest mortality rate was 140 deaths/100,000.

@@

The main outcomes of interest for TB incidence, mortality, and prevalence were plotted using the `geom_point()` function as shown below.
```{r}
p6 <- highburden %>% 
  filter(dimension %in% "Sex") %>%
  filter(indicator_abbr == "mortality") %>% 
  select(c(1,2,3,4,5,6,13)) %>% 
  ggplot(aes(x=fct_reorder(
  setting, `TB mortality (deaths per 100 000 population)`),
  y = `TB mortality (deaths per 100 000 population)`,
  color = subgroup)) +
    geom_point()+
  theme(axis.text.x = element_text(angle = 90))+
  scale_fill_brewer(palette = "Spectral")+
  xlab("setting")
plot(p6) #plots mortality by country from lowest mortality to highest, colored by subgroup
```

@@

## Basic statistical analysis
*(Part 3 of project: linear model fitting and predictions, more to be added)*
TBA: simple statistical models for associations in data

The linear models summary for TB incidence are listed below.The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1. The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`. Based on the p-value for the subgroup variable, sex has a significant association with TB incidence in both models. Males having higher TB incidence is congruent with the exploratory analysis findings. The full model has smaller coefficients compared to `highburden` model. 
```{r}
readRDS(file = here("results", "model fit tables", "incfittable.rds")) #TB incidence model for high burden countries
readRDS(file = here("results", "model fit tables", "incfittableFD.rds")) #TB incidence model for full data
```
The fitted model summaries below are the TB mortality for high burden countries and full data. For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`. Sex is a statistically significant predictor of TB mortality (p-values <0.05), but this does not consider any other factors. The next step in my analysis would be to rearrange the data set so I can predict the main outcomes using the sociodemographic indicators. 
```{r}
readRDS(file = here("results", "model fit tables", "mortfittable.rds")) #TB mortality model for high burden countries
readRDS(file = here("results", "model fit tables", "mortfittableFD.rds")) #TB mortality model for full data
```
The last outcome model I fitted was TB prevalence for high burden countries and the full data. For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0. The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`. 
```{r}
readRDS(file = here("results", "model fit tables", "prevfittable.rds")) #TB prevalence model for high burden countries
readRDS(file = here("results", "model fit tables", "prevfittableFD.rds")) #TB prevalence model for full data
```
The models can be further explored by including `setting` as a predictor, shown in the prediction model below. The prediction model is basic prediction of TB incidence from subgroup and setting. As an example, I predicted the incidence for females and males in India (shown in the plot). The estimates for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) are shown using `geom_point()` and `geom_errorbar()`. I need to explore the prediction modelling further before attempting to predict my main outcomes with more variables. 
```{r}
#prediction model for TB incidence with subgroup and setting as predictors
predinclm_fit <- lm_mod %>% 
  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup + setting,
      data = highburden) #fit linear model
predincfittable <- tidy(predinclm_fit) 
print(predincfittable)

#prediction plot using the above model estimating TB incidence for females and males in India
readRDS(file = here("results", "model fit tables", "prediction1.rds")) 
```
## Full analysis
TBA: final analysis with statistic/ML methods, meaningful figures/tables, load saved products here

@@

## Summary and Interpretation
TBA: summary of findings and why it matters

## Strengths and Limitations
TBA: strengths and limitations of analysis

## Conclusions
TBA: take away messages, citations from bibtex

>>>>

Tuberculosis is an infectious disease caused by Mycobacterium tuberculosis and is the leading infectious cause of death in the world even though the disease is both treatable and curable[@WHO2022]. Approximately 10 million people are infected each year, and 1.5 million die from the disease [@WHO2022]. Humanity has been impacted by TB for centuries, but the burden of disease is not equal across countries and populations [@WHO2022b]. The World Health Organization marked 30 countries as having highest burden of TB due to the number of cases, presence of multi-drug resistance strains, and the high mortality within certain populations such as HIV [@WHO2021b]. To understand more about why some countries or groups of people experience the higher burden, the World Health Organization gathered data on several health equity indicators subgroups. This project will explore and analyze the disparities within the indicators and the effect on tuberculosis outcomes.

@@

The World Health Organization collected data on the inequity surrounding tuberculosis, HIV, and malaria for the [State of inequality report](https://www.who.int/data/inequality-monitor/publications/report_2021_hiv_tb_malaria), and I will be exploring the dataset for TB [@WHO2021]. More information about the data can be found [here](https://heatrepository.blob.core.windows.net/documents/data-repository-indicator-list.pdf?sp=r&st=2022-06-07T14:16:36Z&se=2023-12-30T23:16:36Z&spr=https&sv=2020-08-04&sr=b&sig=4kzThU1QDo55UOQyhWcUd8rPWJ9LxUZdRRI3zl6wKNs%3D) under "Tuberculosis Indicators" [@WHOTBData]. The dataset can be found [here](https://www.who.int/data/inequality-monitor/data#PageContent_C158_Col00) under "Tuberculosis Indicators" [@WHOTBData].

For the tuberculosis dataset, the data was collected from the WHO Global TB programme, TB prevalence surveys, country-specific TB programmes, the WHO Health Equity Monitor database, TB patient cost surveys, and other sources [@WHO2021]. The WHO organized the dataset to be used with the Health Equity Assessment Toolkit which is the built in data analysis and exploration tool. The database contains 10 variables regarding the burden, detection, prevention, knowledge, and social protection and observations for 194 countries over various years. The observations for each variable is further separated by up 7 inequality domains. Not all countries have data available for each year or for each inequality domain. The dataset contains a total of 7473 observations.

@@

Research question: Does high disparities between inequality measures show high correlation to TB incidence, prevalence, and mortality of a country? How do these relationships differ for high burden countries compared to all countries?

@@

@@

## Data aquisition, import, and cleaning

The WHO dataset can be found [here](o%20https://www.who.int/data/inequality-monitor/data#PageContent_C158_Col00) under "Tuberculosis Indicators". This data was downloaded and stored into the `raw_data` folder of the repository as `rawdata`. Information on the indicators, social determinants dimensions, and subgroups can be found in the `raw_data` folder under `202206-metadata-tb.pdf`

```{r load data}
#path to data
data_location <- here::here("data","raw_data","202206-repository-tb.xlsx")

@@

str(rawdata)
```

Please see `processingfile.qmd` for full code describing data importing and cleaning.

@@

For the main part of the data cleaning, I used `pivot_wider()` to allow each indicator to have its own column, and subset the high burden countries (as indicated by the WHO) into a new object called `highburden`. I also checked each indicator using `filter()`, `select()`, and `summary()` functions. These methods will allow for deeper exploration of each indicator in order to explore trends and patterns for each subgroup. An example of the data cleaning is shown below.

```{r}
wide_data <- rawdata %>% 
  select(c(setting, year, indicator_name,
           indicator_abbr, dimension, subgroup,
           estimate, population)) %>% 
  pivot_wider(names_from = "indicator_name", values_from = "estimate")
head(wide_data)

@@

The processed and cleaned data was then saved into the `processed_data` folder under `processeddata.rda` to be used during the exploratory analysis phase. I have loaded the processed data below.

```{r}
#path to data
data_location <- here::here("data","processed_data","processeddata.rda")
#load data. 
load(data_location)
```

## Statistical analysis

For the statistical analysis of the data, I created linear models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator. An example is shown below. This method was repeated for the `wide_data` and `highburden` countries.

@@

Other indicators were also analyzed and fit to a linear model as shown below.

```{r}
# fit linear model with BCG coverage as outcome, subgroups as predictors

@@

## Exploratory/Descriptive analysis

*(Part 2 of project: exploratory analysis)* Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.

@@

#mortality summary and save to file location
summary_mort <- skimr::skim(wide_data$`TB mortality (deaths per 100 000 population)`)
print(summary_mort)
```

The population for most variables were relatively low and skewed to the right. The TB incidence and mortality indicators also showed a skewed distribution. The highest incidence rate was 908 cases/100,000 and highest mortality rate was 140 deaths/100,000.

@@

The main outcomes of interest for TB incidence, mortality, and prevalence were plotted using the `geom_point()` function as shown below.

```{r}
p6 <- highburden %>% 
  filter(dimension %in% "Sex") %>%
  filter(indicator_abbr == "mortality") %>% 
  select(c(1,2,3,4,5,6,13)) %>% 
  ggplot(aes(x=fct_reorder(
  setting, `TB mortality (deaths per 100 000 population)`),
  y = `TB mortality (deaths per 100 000 population)`,
  color = subgroup)) +
    geom_point()+
  theme(axis.text.x = element_text(angle = 90))+
  scale_fill_brewer(palette = "Spectral")+
  xlab("setting")
plot(p6) #plots mortality by country from lowest mortality to highest, colored by subgroup
```

@@

## Basic statistical analysis

*(Part 3 of project: linear model fitting and predictions, more to be added)* TBA: simple statistical models for associations in data

The linear models summary for TB incidence are listed below.The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1. The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`. Based on the p-value for the subgroup variable, sex has a significant association with TB incidence in both models. Males having higher TB incidence is congruent with the exploratory analysis findings. The full model has smaller coefficients compared to `highburden` model.

```{r}
readRDS(file = here("results", "model fit tables", "incfittable.rds")) #TB incidence model for high burden countries
readRDS(file = here("results", "model fit tables", "incfittableFD.rds")) #TB incidence model for full data
```

The fitted model summaries below are the TB mortality for high burden countries and full data. For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`. Sex is a statistically significant predictor of TB mortality (p-values \<0.05), but this does not consider any other factors. The next step in my analysis would be to rearrange the data set so I can predict the main outcomes using the sociodemographic indicators.

```{r}
readRDS(file = here("results", "model fit tables", "mortfittable.rds")) #TB mortality model for high burden countries
readRDS(file = here("results", "model fit tables", "mortfittableFD.rds")) #TB mortality model for full data
```

The last outcome model I fitted was TB prevalence for high burden countries and the full data. For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0. The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`.

```{r}
readRDS(file = here("results", "model fit tables", "prevfittable.rds")) #TB prevalence model for high burden countries
readRDS(file = here("results", "model fit tables", "prevfittableFD.rds")) #TB prevalence model for full data
```

The models can be further explored by including `setting` as a predictor, shown in the prediction model below. The prediction model is basic prediction of TB incidence from subgroup and setting. As an example, I predicted the incidence for females and males in India (shown in the plot). The estimates for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) are shown using `geom_point()` and `geom_errorbar()`. I need to explore the prediction modelling further before attempting to predict my main outcomes with more variables.

```{r}
#prediction model for TB incidence with subgroup and setting as predictors
predinclm_fit <- lm_mod %>% 
  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup + setting,
      data = highburden) #fit linear model
predincfittable <- tidy(predinclm_fit) 
print(predincfittable)

#prediction plot using the above model estimating TB incidence for females and males in India
readRDS(file = here("results", "model fit tables", "prediction1.rds")) 
```

## Full analysis

TBA: final analysis with statistic/ML methods, meaningful figures/tables, load saved products here

@@

## Summary and Interpretation

TBA: summary of findings and why it matters

## Strengths and Limitations

TBA: strengths and limitations of analysis

## Conclusions

TBA: take away messages, citations from bibtex
